% !TEX root = thesis.tex

Just as a thesis would be nothing without the graduate student, a graduate student would be nothing without the support of surrounding friends, family and collaborators. My five years have been blessed with a tremendous collection of all three.

To begin, I thank my parents for their constant support. I marvel at their ability to produce the perfect words of encouragement in good times and bad, brightening the joyful days and smoothing over rocky patches along the path. I cannot imagine navigating this journey without your love and guidance behind me, thank you both for all you have done. And thanks to my brothers, Lucas and Jacob, for being most excellent. I cherish the time we spent together, whether in New York or Toronto.

Next, I thank my advisor, Professor Kevrekidis, who has had the most direct influence on the work that follows. I am grateful for the generosity and compassion he displayed throughout our time together, from the many wonderful meals hosted at ``The Spaniard's'', to his emphasis of family over work, encouraging us to return home during difficult times. I will try to emulate his empathy in my future positions.

Continuing in the academic direction, my collaborator and friend Antonios Zagaris is directly responsible for much of Chapter~\ref{ch:params} and I am very thankful for his input over the past two years. He has been a faithful mentor and source of encouragement, and I have enjoyed equally our discussions on the origins of sloppy models as those on the origins of punk rock. Mahdi Kooshkbaghi was also a great help in the final stages of the project, and I regret that our time together was so short. In just six months he became a close friend, introducing me to proper black tea and Kayhan Kalhor. I will miss them both.

My companions in room ACE44 have ensured that the typical day of research was supplemented with DMAPS jokes and off-topic soccer conversations. From the post-docs: Matt Williams taught me good C programming and talked me into my first 5k. Minseok Choi significantly improved my tennis forehand and showed me the best Korean fried-chicken restaurants around. And Juan Bello-Rivas provided excellent Python packages and had many fruitful suggestions on particular aspects of my work. Of the slew of German visiting students: Robert H\"{o}lzel rekindled my love for chess, Felix Kemeth instituted group lunch days, and Felix Dietrich played an instrumental role in designing the lab coffee grinder. From the collection of fellow graduate students: Karthik Rajendran introduced me to many fundamental topics in our group, Carmeline D'Silva helped me navigate the middle of my graduate student career, Dave Sroczynski was a source of fun both inside and outside of the office, Wendy Jiang kindly obliged the occasional conversation in broken Chinese, and I enjoyed discussing the merits of wood versus bamboo cutting boards with Thomas Thiem.

Outside of academics I was fortunate to enjoy time with a close group of classmates. Dmitry Pozharskiy not only worked across from me for four years, but was also an exellent room-mate throughout. Despite my inability to clean any sort of dish in a reasonable timeframe, I could always depend on his support in matters both professional and private; many hours were passed simply chatting on the sidewalk outside our apartment. Tom Bertalan was an office-mate and friend who would never hesitate to offer his help, be it programming advice or a ride to the airport. Yogesh Goyal and I shared many Indian dinners together, and I appreciate the genuine encouragment he has provided over the years. Will Mulhearn was the easy-going and flexible host of many of our evenings who always welcomed new adventures, including a particularly patriotic roadtrip to Canada. Yile Gu brightened the atmosphere of all our events with his cheerfulness, and was the source of our group's most memorable quotes. Together, the six of us travelled across Europe, organized frequent summer barbeques, and played many hours of board games; I am so grateful for their friendship.

There were numerous others whose company I am thankful for. Zach Domach for being my SCUBA buddy and for the many orchestra outtings, Logan Matthews for many good discussions on religion, Ting Chen for being a caring friend and motivated workout buddy, Elia Altabet for many evenings of conversation over whiskey, Clark Chen and Jack Lu for introducing me to the joys of hot-pot and Rai Rai Ramen, Mike Siedlik for consistently attending our badminton games, and Granton Jindal for fun games of tennis, ping pong and soccer.

To those above and the many others who so enriched the past five years: thank you.